---
title: "Cleaning"
output: html_document
date: "2024-09-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

# Load the data
```{r data}
current_folder <- getwd()

# Load datasets
athletes_raw <- read.csv(file.path(current_folder, "data/athlete_events.csv"))
noc <- read.csv(file.path(current_folder, "data/noc_regions.csv"))
# Merge athletes with noc (region names)
athletes <- merge(athletes_raw, noc, by = "NOC") %>%
  select('ID','Sex','Team','Age','Weight','Height','NOC','Year','Season','Sport','Event','Medal','region') %>% 
  rename_at('region', ~'AthleteCountry') %>%
  # Add new feature WeightHeight
  mutate(WeightHeightRation = Weight/Height)

# Olympic games
games <- read.csv(file.path(current_folder, "data/olympicgames.csv")) %>%
  select('Country','Year')

# For now, to simplify lets use 2016 GPD and Population
countries_gdp <-read.csv(file.path(current_folder, "data/countries_gdp_per_capita.csv")) %>%
  select('CountryName','X2016') %>%
  rename_at('X2016', ~'GDP') %>%
  mutate(GDP = as.integer(GDP)) %>%
  mutate(GDP = GDP/1000000)

countries_pop <- read.csv(file.path(current_folder, "data/countries_pop.csv")) %>% 
  select('CountryName','X2016') %>%
  rename_at('X2016', ~'Population') %>%
  mutate(Population = Population/1000000)

world_data <- read.csv(file.path(current_folder, "data/world-data-2023.csv")) %>%
  select(
    Country,
    Density_P_Km2 = Density..P.Km2.,
    Agricultural_Land_Percent = Agricultural.Land....,
    Land_Area_Km2 = Land.Area.Km2.,
    GDP,
    Population,
    Urban_Population = Urban_population,
    Life_Expectancy = Life.expectancy,
    Gross_Tertiary_Education_Enrollment_Percent = Gross.tertiary.education.enrollment....,
    Physicians_Per_Thousand = Physicians.per.thousand,
    Fertility_Rate = Fertility.Rate,
    Infant_Mortality = Infant.mortality,
    Out_of_Pocket_Health_Expenditure_Percent = Out.of.pocket.health.expenditure,
    Latitude,
    Longitude,
    Unemployment_Rate = Unemployment.rate
  )

head(athletes)
head(games)
head(countries_gdp)
head(countries_pop)
head(world_data)
```
# Data Consistency

First, we want to use the same country nomenclature in all datasets. Let's look for inconsistencies e.g. USA instead of U.S.A or United States.

```{r}
# Clean country data in games table
setdiff(games$Country, athletes$AthleteCountry)

# Same name for different countries
country_mapping <- setNames(c('USA', 'UK', 'Germany', 'Russia', 'Serbia'),c('United States', 'United Kingdom', 'FR Germany', 'Soviet Union', 'Yugoslavia'))
games <- games %>%
  mutate(Country = ifelse(Country %in% names(country_mapping), country_mapping[as.character(Country)], Country))
setdiff(games$Country, athletes$AthleteCountry)
# I no longer have differences

# Clean country data in world_data
setdiff(athletes$AthleteCountry, world_data$Country)
country_mapping <- setNames(c('USA', 'UK', 'Russia'),c('United States', 'United Kingdom', 'Russian Federation'))
countries_pop <- countries_pop %>%
  mutate(CountryName = ifelse(CountryName %in% names(country_mapping), country_mapping[as.character(CountryName)], CountryName))
setdiff(athletes$AthleteCountry, countries_pop$CountryName)
# I no longer have differences
```

From 1992 to 2016, we have Olympics every 2 years. Separation of Summer and Winter Games decision.
For simplifying purposes, lets group summer and winter Olympics into the same year:
```{r}
athletes %>% distinct(Year) %>% arrange(Year) 
year_mapping <- setNames(c(1992, 1996, 2000, 2004, 2008, 2012), c(1994, 1998, 2002, 2006, 2010, 2014))
athletes <- athletes %>%
  mutate(AdjustedYear = ifelse(Year %in% names(year_mapping), year_mapping[as.character(Year)], Year))
games <- games %>%
  mutate(AdjustedYear = ifelse(Year %in% names(year_mapping), year_mapping[as.character(Year)], Year))
```


# Null values
```{r}
null_count <- sapply(athletes, function(x) sum(is.na(x)))
total_rows <- nrow(athletes)
data.frame(Column = names(null_count), 
           NullCount = null_count, 
           NullPercentage = (null_count / total_rows) * 100)
```
We are missing 3% of the *Age* column values, 23 and 22% of values in *Weight* and *Height* columns respectively.

In the latest years, we have much more data about Weight and Height going from ~90% to ~0% missing values:
```{r}
#athletes %>% filter(Year > 1990) %>%
  #group_by(AthleteCountry) %>% 
  #summarize(ssumnull = sum(is.na(Height)), .groups='drop') %>% 
  #arrange(desc(ssumnull))
  
```

# Data Types
```{r}
world_data <- world_data %>%
    mutate(
    Density_P_Km2 = as.numeric(gsub(",", "", Density_P_Km2)),
    Agricultural_Land_Percent = as.numeric(gsub("%", "", Agricultural_Land_Percent)) / 100,
    Land_Area_Km2 = as.numeric(gsub(",", "", Land_Area_Km2)),
    GDP = as.numeric(gsub("[^0-9.]", "", GDP)),
    Population = as.numeric(gsub(",", "", Population)),
    Urban_Population = as.numeric(gsub(",", "", Urban_Population)),
    Gross_Tertiary_Education_Enrollment_Percent = as.numeric(gsub("%", "", Gross_Tertiary_Education_Enrollment_Percent)) / 100,
    Out_of_Pocket_Health_Expenditure_Percent = as.numeric(gsub("%", "", Out_of_Pocket_Health_Expenditure_Percent)) / 100,
    Unemployment_Rate = as.numeric(gsub("%", "", Unemployment_Rate)) / 100
  )

write.csv(world_data, file.path(current_folder, "data/clean_world_data.csv"), row.names = FALSE)
```

# Merge data and Save
```{r}
clean_data <- merge(athletes %>% select(-Year), games, by = "AdjustedYear") %>%
  mutate(HomeAway = ifelse(Country == AthleteCountry, "Home", "Away"))

current_folder <- getwd()
write.csv(clean_data, file.path(current_folder, "data/clean_data.csv"), row.names = FALSE)
```

# Create new columns
- HomeAway
- GoldMedal
- Medal