---
title: "EDA"
output: html_document
date: "2024-09-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

## EDA

# Load the data
```{r data}
current_folder <- getwd()
clean_data <- read.csv(file.path(current_folder, "data/clean_data.csv"))

head(clean_data)
```

## EDA

We have data from 1896 to 2016:
```{r}
# YEAR -----------------------------------------------
c(min(clean_data$AdjustedYear), max(clean_data$AdjustedYear))
```
The fist summer Summer Olympics (1896) included 9 sports, the last (2016) ones include 34.
The fist summer Winter Olympics (1924) included 10 sports, the last (2014) ones include 15.

# Sport
Q1: How many sports in each season? And how the #sports per season changed with time?
```{r sports, message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Season,Year) %>% summarise(n_distinct(Sport), .groups='drop') %>%
  group_by(Season) %>% 
  filter(Year==max(Year) | Year==min(Year))

clean_data %>%
  group_by(Year, Season) %>%
  summarize(NSports = n_distinct(Sport), .groups='drop') %>%
  ggplot(aes(x = Year, y = NSports, color=Season)) +
  geom_line() +
  geom_point() +
  labs(title = "Distinct Sports by Year by Season",
       x = "Year",
       y = "Number of Distinct Sports") +
  theme_minimal()
```

Q2: Which sports have been in the Olympic games every single year?
```{r}
clean_data %>%
  group_by(Sport) %>%
  summarise(Years_Present = n_distinct(AdjustedYear)) %>%
  filter(Years_Present == n_distinct(clean_data$AdjustedYear)) %>%
  pull(Sport)
```

Q3: Which sports have been played just once in the Olympics?
```{r}
clean_data %>%
  group_by(Sport) %>%
  summarize(YearCount = n_distinct(Year),
            YearPlayed = first(Year[which(n_distinct(Year) == 1)])) %>%
  filter(YearCount == 1)
```

## Height and Weight
Q1: Is there high correlation?
```{r}
cor(clean_data$Weight, clean_data$Height, use = "complete.obs")
```

Q2: Is the height distribution of athletes different in different sports?
```{r}
clean_data %>% 
  filter(Sport %in% c('Basketball','Football')) %>% 
  ggplot(aes(x = Height)) +
  geom_histogram(binwidth = 5, fill = 'blue', color = 'black', alpha = 0.7) +
  facet_wrap(~ Sport) +
  theme_minimal() +
  labs(title = 'Height Distribution by Sport', x = 'Height (cm)', y = 'Frequency')
```

Q3: T test to check if Basketball players are taller than Football players
```{r}
basketball_players <- clean_data %>% filter(Sport == "Basketball")
football_players <- clean_data %>% filter(Sport == "Football")
t.test(basketball_players$Height, football_players$Height, alternate='greater')
```

Q4: ANOVA Test: Is the weight distribution of athletes different in different sports?
```{r}
clean_data %>% 
  filter(Sport %in% c('Wrestling','Weightlifting','Judo')) %>% 
  ggplot(aes(x = Weight)) +
  geom_histogram(binwidth = 5, fill = 'blue', color = 'black', alpha = 0.7) +
  facet_wrap(~ Sport) +
  theme_minimal() +
  labs(title = 'Weight Distribution by Sport', x = 'Weight (kg)', y = 'Frequency')

weight_diffs <- clean_data %>% 
  filter(Sport %in% c('Wrestling','Weightlifting','Judo'))
anova <- aov(Weight~Sport, data=weight_diffs)
summary(anova)

TukeyHSD(anova, conf.level = 0.95)
plot(TukeyHSD(anova, conf.level = 0.95))
```

Q5: Has weight and height changed over time?
Which sports have seen largest differences in weight over time?
```{r}
clean_data %>%
  group_by(AdjustedYear) %>%
  summarise(
    Mean_Height = mean(Height, na.rm=T),
    Mean_Weight = mean(Weight, na.rm=T)
  ) %>%
  pivot_longer(cols = c(Mean_Height, Mean_Weight), names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = AdjustedYear, y = Value)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Metric, scales = "free_y", ncol = 1) +
  theme_minimal() +
  labs(title = 'Mean Height and Weight Over Years by Sport', x = 'Year', y = 'Value')

# Q5: Which sports have seen largest differences in weight over time?
clean_data %>%  
  group_by(Sport) %>%
  summarise(
    Mean_Height = mean(Height, na.rm=T),
    SD_Height = sd(Height, na.rm=T),
    Mean_Weight = mean(Weight, na.rm=T),
    SD_Weight = sd(Weight, na.rm=T)
    ) %>%
  arrange(desc(SD_Height))
# Basketball has seen larger changes in the height of the athletes over time
# Plot
clean_data %>%
  filter(Sport %in% c('Basketball')) %>%
  group_by(AdjustedYear,Sport) %>%
  summarise( Mean_Height = mean(Height, na.rm=T) ) %>%
  ggplot(aes(x = AdjustedYear, y = Mean_Height, color=Sport)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = 'Mean Height Over Years by Sport', x = 'Year', y = 'Mean Weight (cm)')
```


Q6: Which sports have seen largest differences in weight over time?
```{r}
clean_data %>%  
  group_by(Sport) %>%
  summarise(
    Mean_Weight = mean(Weight, na.rm=T),
    SD_Weight = sd(Weight, na.rm=T)
  ) %>%
  arrange(desc(SD_Weight))
# Weightlifting, Judo and Wrestling have seen larger changes in the weight of the athletes over time
# Plot
clean_data %>%
  filter(Sport %in% c('Weightlifting','Judo','Wrestling')) %>%
  group_by(AdjustedYear,Sport) %>%
  summarise( Mean_Weight = mean(Weight, na.rm=T) ) %>%
  ggplot(aes(x = AdjustedYear, y = Mean_Weight, color=Sport)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = 'Mean Weight Over Years by Sport', x = 'Year', y = 'Mean Weight (Kg)')
```

## Country

Q1: Has country participation increased with time?
```{r}
clean_data %>%
  group_by(Year, Season) %>%
  summarize(NCountries = n_distinct(AthleteCountry), .groups='drop') %>%
  ggplot(aes(x = Year, y = NCountries, color=Season)) +
  geom_line() +
  geom_point() +
  labs(title = "Distinct Countries by Year",
       x = "Year",
       y = "Number of Distinct Countries") +
  theme_minimal()

clean_data %>%
  group_by(Year, Season) %>%
  summarize(NAthletes = n_distinct(ID), .groups='drop') %>%
  ggplot(aes(x = Year, y = NAthletes, color=Season)) +
  geom_line() +
  geom_point() +
  labs(title = "#Athlethes by Year",
       x = "Year",
       y = "Number of Distinct Athletes by Season") +
  theme_minimal()
```

Country (and therefore #athletes) participation in the Olympic Games has increased significantly in the last 20 years
Higher participation in the Summer games compared to the Winter ones
A dip in the Summer Olympic Games in 1980 - Boycott of the Moscow Olympics by US to protest against Afghanistan invasion.

Q2: Which sports offer the more medals (gold medals) on average?
```{r}
clean_data %>% 
  filter(Year == 2016) %>%
  group_by(Sport) %>% summarize(NMedals = sum(!is.na(Medal) & Medal=='Gold')) %>% arrange(desc(NMedals)) %>%
  ggplot(aes(x = reorder(Sport, NMedals), y = NMedals)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = NMedals), size = 3, color = "black") +
  coord_flip() +
  labs(title = "Number of Medals by Sport",
       x = "Sport",
       y = "Number of Medals") +
  theme_minimal()

clean_data %>% 
  filter(Year == 2016 & !(is.na(Medal))) %>%
  group_by(Sport, Medal) %>%
  summarize(NMedals = sum(!is.na(Medal)), .groups = 'drop') %>%
  mutate(Medal = factor(Medal, levels = c("Gold", "Silver", "Bronze"))) %>%
  ggplot(aes(x = reorder(Sport, NMedals), y = NMedals, fill = Medal)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = NMedals), position = position_stack(vjust = 0.5), size = 3, color = "white") +
  coord_flip() +
  labs(title = "Number of Medals by Sport (2016)",
       x = "Sport",
       y = "Number of Medals") +
  scale_fill_manual(values = c("Gold" = "gold", "Silver" = "grey", "Bronze" = "brown")) +
  theme_minimal()
```
Why are there more gold medals than silver or bronze in some sports e.g. athletics?
```{r}
clean_data %>% 
  filter(Sport == 'Athletics' & Year == 2016 & !(is.na(Medal))) %>%
  group_by(Event, Medal) %>%
  summarize(NMedals = sum(!is.na(Medal)), .groups = 'drop') %>%
  mutate(Medal = factor(Medal, levels = c("Gold", "Silver", "Bronze"))) %>%
  ggplot(aes(x = reorder(Event, NMedals), y = NMedals, fill = Medal)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = NMedals), position = position_stack(vjust = 0.5), size = 3, color = "white") +
  coord_flip() +
  labs(title = "Number of Medals by Sport (2016)",
       x = "Sport",
       y = "Number of Medals") +
  scale_fill_manual(values = c("Gold" = "gold", "Silver" = "grey", "Bronze" = "brown")) +
  theme_minimal()

clean_data %>% 
  filter(Event == "Athletics Women's 4 x 400 metres Relay" & Year == 2016 & Medal == 'Gold')
```

# Olympics Host Advange
Is there a significant advantage for countries playing at home vs away?
```{r}
# Get the list of countries that have hosted
host_countries_lst <- unique(clean_data$Country)

# Loop through the list of host countries
for (country in host_countries_lst) {
  
  # Filter data for the current country
  medals_per_country <- clean_data %>%
    group_by(AdjustedYear, AthleteCountry) %>%
    summarise(Medals = sum(!is.na(Medal)), .groups = 'drop')
  
  hosting_years <- clean_data %>%
    filter(HomeAway == 'Home' & Country == country & Season == 'Summer') %>%
    pull(AdjustedYear) %>%
    unique()
  
  plot_data <- clean_data %>% 
    group_by(AdjustedYear) %>% 
    summarize(TotalMedals = sum(!is.na(Medal)), .groups = 'drop') %>%
    left_join(medals_per_country, by = 'AdjustedYear') %>%
    mutate(MedalsPercent = Medals / TotalMedals) %>%
    filter(AthleteCountry == country)
  
  country_medals <- plot_data %>% filter(AdjustedYear %in% hosting_years)
  
  # Create the plot
  p <- ggplot(data = plot_data, aes(x = AdjustedYear)) +
    geom_bar(data = country_medals, aes(y = MedalsPercent), stat = "identity", width = 4, fill = "skyblue", alpha = 0.7) +
    geom_line(aes(y = MedalsPercent)) +
    labs(title = paste("Percentage of Medals Won Over Time -", country),
         x = "Year",
         y = "Percentage of Medals") +
    theme_minimal()
  
  # Print the plot for the current country
  print(p)
}
```

# Correlation of medals to GDP or Population
